A mechanization of semi-simplicial and semi-cubical sets in the calculus of inductive constructions

### 2022-10-10 paper/paper.tex
paper: MAJ

% Truncated sets, otherwise referred to as X: the general macro
\newcommandx{\X}[3][1,2,3]{
  \ensuremath{\mathsf{\color{carolina}{X}}_{#1}^{#2}}
  \setkeys*[bonak]{X}{#3}
}

@@

% Universe, referred to as U
\newcommand{\U}[1][]{\ensuremath{\mathsf{\color{spanish-blue}{HSet}}_{#1}}}

@@

% These are renewed in tables, but let's define them now for in-document
% typesetting
\newcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\newcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\newcommandx{\filler}[2][1,2]{\prim{filler}[][#1][#2][][]}
\newcommandx{\restrf}[1][1]{\restr{frame}[][#1][][][][][]}

@@

The outline of the paper is as follows. First, we briefly describe semi-simplicial and semi-cubical sets, as it appears in mathematical literature, and our generalization to $\nu$-sets. We then proceed to explain our construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning in sets.

@@

\subsection{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the points are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over a singleton set of colors.

@@

\begin{example}[$\DeltaPlus^1$]
  In dimension $1$, the standard augmented semi-simplex can be pictured as a point, where the unique color is kept implicit; the unique morphism in $\Hom(0, 1)$ is not pictured. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$:

@@

\begin{example}[$\DeltaPlus^2$]
  In dimension $2$, the standard augmented semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. Again, the unique morphim in $\Hom(0, 2)$ is not pictured.

@@

\subsection{Iterated parametricity}
One step of parametricity can be seen as interpreting a point by a line. Iterating parametricity shall lead to map terms typed in a context with $n$ declarations of an axis as $n$-fillers. We shall interpret a typing derivation with $n$ axis declarations as an $n$-filler of typing derivations in the target. For instance, a typing derivation with exactly one declaration of an axis will be interpreted by 3 typing derivations in the target, two of them typing points and the third one typing a proof, seen as a line, together with the fact that these two points are related (as in ordinary parametricity), while a typing derivation with $n$ declarations of an axis shall be interpreted as a $3^n$ derivations, split into $2^n$ derivations of points.

@@

and so on. Thus, we need first to make the definition of $\filler[n][p]$ dependent over the part of the full filler built up to this point and secondly to accumulate this information. At the end of the accumulation of the process, i.e. at the time of building $\filler[n+1][n]$, a complete $n$-frame will have been built. So, let us call $\filler[n][p]$ the partial $n$-frame built up to the stage $p$ of the construction. Thus, we refine our definition into something like:

@@

\section{Type-theoretic construction of \texorpdfstring{$\nu$}{nu}-sets}
% Abbreviated tables in this section
%
% The rule for implicit arguments for each table:
% show 1 explicit in first two parts
% show E + 1 explicit in the third part
\appendmask[bonak]{layer}[D]
\appendmask[bonak]{filler}[D]
\appendmask[bonak]{restrframe}[D]
\appendmask[bonak]{restrlayer}[D, d]
\appendmask[bonak]{restrfiller}[D, d]
\appendmask[bonak]{cohframe}[D]
\appendmask[bonak]{cohlayer}[D, d]
\appendmask[bonak]{cohfiller}[D, d]

@@

% For just the second table, keep the universe letter in fullframe
\newcommandx{\fullframe}[3][1,2,3]{\prim{fullframe}[#1][#2][][][#3]}

>>>>

% Truncated sets, otherwise referred to as X: the general macro
\newcommandx{\X}[3][1,2,3]{
  \ensuremath{{\color{carolina}{\nu\textsf{Set}}}_{#1}^{#2}}
  \setkeys*[bonak]{X}{#3}
}

@@

% Universe, referred to as U
\newcommandx{\U}[1][1=]{\ensuremath{\mathsf{\color{spanish-blue}{HSet}}_{#1}}}

@@

% These are renewed in tables, but let's define them now for in-document
% typesetting
\newcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\newcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\newcommandx{\filler}[2][1,2]{\prim{filler}[][#1][#2][][]}
\newcommandx{\restrf}[1][1]{\restr{frame}[][#1][][][][][]}
\newcommandx{\fullframe}[1][1]{\prim{fullframe}[][#1][][][]}

@@

The outline of the paper is as follows. First, we briefly describe semi-simplicial and semi-cubical sets, and our generalization to $\nu$-sets. We then proceed to explain our construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning in sets.

@@

\subsection{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the connected components are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over a singleton set of colors.

@@

\begin{example}[$\DeltaPlus^1$]
  In dimension $1$, the standard augmented semi-simplex can be pictured as a point, colored black, corresponding to the unique morphism in $\Hom(0, 1)$. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$:

@@

\begin{example}[$\DeltaPlus^2$]
  In dimension $2$, the standard augmented semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use the color black to denote the unique morphims in $\Hom(0, 1)$ and $\Hom(0, 2)$.

@@

\subsection{Intuition for our formal construction}
To assign $X_0$, $X_1$, $X_2$, $\ldots$ in the above indexed representation types, we need to exhibit a recursive definition over our building blocks, which we will call $\framep$, $\layer$, and $\filler$. As a first step, we assign every $X_n$ the type $\fullframe[n] \rightarrow \U$, which is a form of uncurrification (i.e. the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$). The question is then to find the recursive recipe defining $\fullframe[n]$. First, $\fullframe[n]$ depends on all $X_i$ upto $n - 1$. So, we need to package up $X_i$, for $i < n$, into a type, which we call $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. For $D: \X[][<n][]$, the initial prefix of $\X[][0][]$ to $\X[][n - 1][]$, $\X[][n][]$ has type $\fullframe[n](D) \rightarrow \U$. We shall give the full definition of $\fullframe[n]$ shortly, but first assuming $\fullframe[n]$ is defined, we are able to take the coinductive limit of the $\X[][<n][]$, as shown in \ref{tab:coind}, therefore fully characterizing the indexed definition of $\nu$-sets.

@@

and so on. Thus, we need first to make the definition of $\filler[n][p]$ dependent over the part of the full filler built up to this point and secondly to accumulate this information. At the end of the accumulation of the process, i.e. at the time of building $\filler[n+1][n]$, a complete $n$-frame will have been built. So, let us call $\framep[n][p]$ the partial $n$-frame built up to the stage $p$ of the construction. Thus, we refine our definition into something like:

@@

\subsection{Formal construction of \texorpdfstring{$\nu$}{nu}-sets}
% Abbreviated tables in this section
%
% The rule for implicit arguments for each table:
% show 1 explicit in first two parts
% show E + 1 explicit in the third part
\appendmask[bonak]{layer}[D]
\appendmask[bonak]{filler}[D]
\appendmask[bonak]{restrframe}[D]
\appendmask[bonak]{restrlayer}[D, d]
\appendmask[bonak]{restrfiller}[D, d]
\appendmask[bonak]{cohframe}[D]
\appendmask[bonak]{cohlayer}[D, d]
\appendmask[bonak]{cohfiller}[D, d]

@@

% For just the second table, keep the universe letter in fullframe
\renewcommandx{\fullframe}[3][1,2,3]{\prim{fullframe}[#1][#2][][][#3]}
